\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-02 paper.qmd
small change, hopefully not a conflict

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the $k-D$ minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

First, a random variable $a$ is generated from a uniform distribution between $-\frac{\pi}{2}$ and $\frac{\pi}{2}$, representing the parameter $a$ is in $T$. Next, three variables $x_1$, $x_2$, and $x_3$ are computed based on $a$ to construct the S-curve data. $x_1$ is calculated as the sine of $a$ (@eq-eqn1), $x_2$ as twice the value of another uniformly distributed random variable between 0 and 1 (@eq-eqn2), and $x_3$ as the cosine of $a$ minus 1, multiplied by the sign of $a$ (@eq-eqn3). Additionally, four noise dimensions are introduced to the dataset to introduce noise variability: $x_4$, $x_5$, $x_6$, and $x_7$, each generated from uniform distributions within specific intervals (@eq-eqn4, @eq-eqn5, and @eq-eqn6 respectively). Finally, these variables are combined to form the S-curve dataset $T$.

$$
x_1 = \sin(3 \pi a)
$$ {#eq-eqn1}

$$
x_2 = 2.0 \times U(0, 1)
$$ {#eq-eqn2}

$$
x_3 = \text{sign}(a) \times (\cos(3 \pi a) - 1)
$$ {#eq-eqn3}

$$
x_4, x_5 \sim U(-0.02, 0.02)
$$ {#eq-eqn4}

$$
x_6 \sim U(-0.1, 0.1)
$$ {#eq-eqn5}

$$
x_7 \sim U(-0.01, 0.01)
$$ {#eq-eqn6}

>>>>

Many new methods for NLDR have emerged in recent years, all designed to better capture specific structures potentially existing in \pD. Here we focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP and PaCMAP. tNSE and UMAP can be considered to produce the \kD minimizing the divergence between two distributions, where the distributions are modeling the inter-point distances. PHATE, TriMAP and PaCMAP are examples of diffusion processes [@coifman2005] spreading to capture geometric shapes, that include both global and local structure.

@@

### 2024-04-02 paper.tex
small change, hopefully not a conflict

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \pD. Here we
focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the \(k-D\)
minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{coifman2005} spreading
to capture geometric shapes, that include both global and local
structure.

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$X$ & $p$-dimensional data\\
$Y$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d-dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$y_{max}$ & maximum value of the scaled second embedding component\\
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
\addlinespace
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
\addlinespace
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@

The S-curve model, denoted as \(T\), is simulated through several steps.
First, a random variable \(x\) is generated from a uniform distribution
between \(-\frac{\pi}{2}\) and \(\frac{\pi}{2}\), representing the
parameter \(x\) in the S-curve model. Next, three variables \(x_1\),
\(x_2\), and \(x_3\) are computed based on \(x\) to construct the
S-curve data. \(x_1\) is calculated as the sine of \(x\)
(@ref(eq:eqn2)), \(x_2\) as twice the value of another uniformly
distributed random variable between 0 and 1, and \(x_3\) as the cosine
of \(x\) minus 1, multiplied by the sign of \(x\). Additionally, four
noise dimensions are introduced to the dataset to introduce variability:
\(x_4\), \(x_5\), \(x_6\), and \(x_7\), each generated from uniform
distributions within specific intervals. Finally, these variables are
combined to form the S-curve dataset \(T\).

\begin{equation} \label{eq-eqn1}
x_1 = \sin(3 \pi x)
\end{equation}

\begin{equation} \label{eq:eqn2}
x_2 = 2.0 \times U(0, 1)
\end{equation}

\begin{equation} \label{eq:eqn3}
x_3 = \text{sign}(x) \times (\cos(3 \pi x) - 1)
\end{equation}

\begin{equation} \label{eq:eqn4}
x_4, x_5 \sim U(-0.02, 0.02)
\end{equation}

\begin{equation} \label{eq:eqn5}
x_6 \sim U(-0.1, 0.1)
\end{equation}

\begin{equation} \label{eq:eqn6}
x_7 \sim U(-0.01, 0.01)
\end{equation}

>>>>

Many new methods for NLDR have emerged in recent years, all designed to
better capture specific structures potentially existing in \pD. Here we
focus on five currently popular techniques, tSNE, UMAP, PHATE, TriMAP
and PaCMAP. tNSE and UMAP can be considered to produce the
\kD minimizing the divergence between two distributions, where the
distributions are modeling the inter-point distances. PHATE, TriMAP and
PaCMAP are examples of diffusion processes \citep{coifman2005} spreading
to capture geometric shapes, that include both global and local
structure.

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$X$ & $p$-dimensional data\\
$Y$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d-dimensional$ linear projection of $p$-dimensional data\\
$y_{max}$ & maximum value of the scaled second embedding component\\
\addlinespace
$r_1$, $r_2$ & range of the first and  second embedding component\\
$h_b$, $w_b$ & height and width of the hexagon\\
$a_r$ & aspect ratio of the 2D layout before scaling\\
$h_r$ & hexagon ratio\\
$s$ & hexagonal size (radius of the outer circle surrounding the hexagon)\\
\addlinespace
$(x_s, \ y_s)$ & starting coordinates of the hexagonal grid\\
$q_x$, $q_y$ & buffer amount along the x and y axes\\
$h_s$, $v_s$ & horizontal and vertical distance between adjacent hexagon bin centroids\\
$d_x$, $d_y$ & horizontal spacing and adjusted vertical spacing for hexagonal coordinates\\
$b_x$, $b_y$ & number of hexagon bins along the x and y axes\\
\addlinespace
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_k$ & number of observations within the $k^{th}$ hexagon\\
$(h^k_{x_i}, h^k_{y_i})$ & hexagonal grid coordinates of the $k^{th}$ hexagon\\
$C^{(j)}$ & $j$-dimensional bin centers\\
\bottomrule
\end{tabular}
\endgroup{}

@@
